---
title: "Singh and Block Wine Grape TAGseq Differential Gene Expression Analysis"
author: "Bioinformatics Core"
output:
  html_document:
    keep_md: true
  pdf_document: default
---

```{r load_packages, echo=FALSE, warning=FALSE, message=FALSE, results='hide', include = FALSE}
#if (!requireNamespace("BiocManager", quietly = TRUE)) {
#  install.packages("BiocManager")
#}

list.of.packages <- c("edgeR", "emmeans", "kableExtra", "knitr", "stringr", "ggplot2", "dplyr", "variancePartition", "readxl")

#if (!all(list.of.packages %in% rownames(installed.packages()))) {
#  BiocManager::install(list.of.packages)
# c}

tmp <- lapply(list.of.packages, require, character.only = TRUE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

```{r}
options(width = 1200)
```


```{r anno, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
# setwd("/home/bnjenner/Center/work/CORE/analysis/Block_D_UCD/2025_April_Singh_Grape_TAGseq/2025_April_Singh_Grape_TAGseq/") # set to your working directory

anno <- read.delim("GCF_030704535.1_ASM3070453v1_and_GRBV_annotation_info.txt")
colnames(anno) <- c("Gene.stable.ID", "DB.xref", "Description", "Molecule.Type")
anno <- anno[!duplicated(anno$Gene.stable.ID),]
colnames(anno)
anno$NCBI <- gsub("GeneID:", "", anno$DB.xref, fixed = TRUE)

#### Adding ensembl IDs
ensembl <- read.delim("NCBI_to_ensembl_114.txt") 
names(ensembl) <- c("Ensembl", "NCBI")
# pick first Ensembl ID for each NCBI ID
ensembl <- group_by(ensembl, NCBI) %>% slice_head(n = 1) %>% ungroup
ensembl$NCBI <- as.character(ensembl$NCBI)
anno <- left_join(anno, ensembl, by = "NCBI")

anno$Description <- gsub("\t", " ", anno$Description)
anno$Description <- gsub("\n", " ", anno$Description)
```

```{r read_count_data, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
counts <- read.delim("./2025_April_Singh_Grape_TAGseq_ASM3070453v1_counts.txt", row.names = 1, sep = "\t")
colnames(counts) <- gsub("^X", "", colnames(counts))
counts <- counts[,order(colnames(counts))]
head(counts)
```

##### Project Metadata
```{r metadata, echo=FALSE, warning=FALSE, message=FALSE}
#mdata <- read.delim("../Sample_Metadata.csv", sep = ",")
mdata <- read_excel("Correct.Sample_Metadata.xlsx")
mdata <- dplyr::rename(mdata, sample = Sample, season = `Sampling Season`, treatment = `Infection Group`, stage = `Maturity stage`, vine = `Vine ID`, ysi = `Years Since Infection`)
mdata <- mdata[order(mdata$sample),]

# Remove Dropped Samples
mdrop <- which(!(mdata$sample %in% colnames(counts)))
mdata <- mdata[-mdrop,]
#mdata$sample == colnames(counts) # Ensure Order

# Remove 2021 Season from Metadata and Counts
mdrop <- which(mdata$season == 2021)
mdata <- mdata[-mdrop,]
counts <- counts[,-mdrop]

treatment <- as.factor(mdata$treatment)
season <- as.factor(mdata$season)
stage <- as.factor(mdata$stage)
ysi <- as.factor(gsub("-", "Uninfected", mdata$ysi))
vine <- as.factor(mdata$vine)
replicate <- as.factor(mdata$replicate)

group <- interaction(treatment, ysi)
group <- gsub("(+)", "", group, fixed = TRUE)
group <- gsub("RB(-)", "Uninfected", group, fixed = TRUE)
group <- factor(group)

metadata <- data.frame("group" = group,
                       "stage" = stage,
                       "vine" = vine,
                       "season" = season,
                       row.names = colnames(counts))
metadata %>%
  kbl(align = 'c', row.names = TRUE) %>%
  add_header_above(c("Metadata" = 5), font_size = "larger", bold = T, align = "l") %>%
  scroll_box(width = "550px", height = "500px") %>% 
  kable_styling()
```


### Library Size
```{r fig1, fig.width=9,fig.height=6, echo=FALSE, warning=FALSE, message=FALSE}
df <- data.frame("Sample" = colnames(counts),
                 "Group" = group,
                 "Count" = colSums(counts))

p1 <- ggplot(df, aes(fill=Group, y=Count, x=Sample)) +
      geom_bar(stat="identity", position="dodge", ) +
      labs(y= "Counts", x = "Sample") + ggtitle("Library Size")
p1
```


```{r, include = F}
# Specify model formula to be used in limma.
form.fixed <- " ~group*stage"
mm <- model.matrix(as.formula(form.fixed), data = metadata)
colnames(mm)
```


```{r model_and_filter, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
d0 <- DGEList(counts)

d0 <- calcNormFactors(d0)
nrow(d0$counts)

logcpm_unfilt <- cpm(d0, prior.count=2, log=TRUE)
colnames(logcpm_unfilt) <- colnames(logcpm_unfilt)

# write.table(logcpm_unfilt,"2025_April_Singh_Grape_TAGseq_ASM3070453v1_2021_normalized_counts.txt", sep="\t", quote=F)
# Make rownames Ensembl IDs
tmp <- as.data.frame(logcpm_unfilt)
tmp$Gene.stable.ID <- rownames(tmp)
tmp <- left_join(tmp, select(anno, Gene.stable.ID, Ensembl), by = "Gene.stable.ID")
tmp <- subset(tmp, Ensembl != "" & !duplicated(Ensembl))
rownames(tmp) <- tmp$Ensembl
tmp <- select(tmp, -Gene.stable.ID, -Ensembl)
write.table(tmp,"2025_April_Singh_Grape_TAGseq_ASM3070453v1_2021_normalized_counts_ENSEMBL.txt", sep="\t", quote=F)

keep <- filterByExpr(d0, mm)
d <- d0[keep,]

na.omit(d)
```


### Number of Expressed Genes

Genes were filtered according to edgeR function "filterByExpr".

```{r expressed, echo=FALSE}
nrow(d)
```

### Visualizaing your data with a Multidimensional scaling (MDS) plot. 

#### Colored by Group
```{r mds1, fig.width=6, fig.height=8, echo=FALSE, warning=FALSE, message=FALSE}
plotMDS(d, col = as.numeric(group), cex=1, plot = T)
```

#### Colored by Stage
```{r mds2, fig.width=6, fig.height=8, echo=FALSE, warning=FALSE, message=FALSE}
plotMDS(d, col = as.numeric(stage), cex=1, plot = T)
```

#### Colored by Timepoint
```{r mds3, fig.width=6, fig.height=8, echo=FALSE, warning=FALSE, message=FALSE}
plotMDS(d, col = as.numeric(ysi), cex=1, plot = T)
```


```{r, include = F}
# Specify model formula to be used in limma.
param <- SnowParam(4, "SOCK", progressbar=TRUE)
form <- ~group*stage + (1|vine)

# estimate weights using linear mixed model of dream
vobjDream <- voomWithDreamWeights(d, form, metadata, BPPARAM=param)

# Use emmeans to get contrasts
z <- rnorm(n = nrow(mm))
dummy.mod <- lm(as.formula(paste0("z", form.fixed)), 
                data = metadata)

# Contrasts between Groups
pairs <- pairs(emmeans(dummy.mod, ~group | stage), reverse = TRUE)
contrast.matrix <- pairs@linfct
tmp <- pairs@grid
rownames(contrast.matrix) <- paste0(tmp$contrast, "_", pairs@grid$stage)
contrasts.group <- rownames(contrast.matrix)
contrast.matrix.group <- contrast.matrix


# Contrasts between stages
pairs <- pairs(emmeans(dummy.mod, ~stage | group), reverse = TRUE)
contrast.matrix <- pairs@linfct
tmp <- pairs@grid
rownames(contrast.matrix) <- paste0(tmp$contrast, "_", pairs@grid$group)
contrasts.stage <- rownames(contrast.matrix)
contrast.matrix.stage <- contrast.matrix

contrast.matrix <- rbind(contrast.matrix.group, contrast.matrix.stage)

L <- t(contrast.matrix)
fitmm <- dream(vobjDream, form, metadata, L)
fitmm <- eBayes(fitmm)
```

## Differential Expression Results

### Output files include the columns:

* logFC: log2 fold change, with the group listed first in the comparison or file name being the numerator of the fold change
* AveExpr: Average expression across all samples, in log2 counts per million reads
* P.Value: Raw p-value from the test that the log fold change differs from 0
* adj.P.Val: Benjamini-Hochberg false discovery rate adjusted p-value

```{r, echo=FALSE, warning=FALSE, message=FALSE}
dir.create(file.path(getwd(), "./DE_Results_2022"))
dir.create(file.path(getwd(), "./DE_Results_2022/BetweenGroups/"))
tables <- lapply(contrasts.group, function(contr) {
  # tmp <- contrasts.fit(fitmm, contrasts = contr)
  # tmp <- eBayes(tmp)
  tmp2 <- topTable(fitmm, coef = contr, n = Inf, sort.by = "P")
  tmp2$Gene.stable.ID <- rownames(tmp2)
  tmp2 <- left_join(tmp2, anno)
  tmp2 <- dplyr::select(tmp2, Gene.stable.ID, logFC, t, AveExpr, P.Value, adj.P.Val, DB.xref, Description, Ensembl, NCBI)
  filename <- gsub(" - ", "_v_", contr)
  filename <- gsub(", ", "_", filename, fixed = T)
  filename <- gsub("+", "", filename, fixed = T)
  filename <- gsub(" ", "_", filename, fixed = T)
  filename <- paste0(filename, "_DE_analysis")
  filename <- paste0(filename, ".txt")
  
  write.table(tmp2, file = paste0("./DE_Results_2022/BetweenGroups/",filename), sep = "\t", quote = F, row.names = F)
  return(list('', nump = length(which(tmp2$adj.P.Val < 0.05))))
})

sigps <- unlist(lapply(tables, function(x)x[[2]]))
sumtab <- data.frame(Comparison = contrasts.group, `Number of genes with adjusted P < 0.05` = sigps,
                     check.names = F)

sumtab %>%
  kbl(align = 'c') %>%
  add_header_above(c("Overview of results" = 2), font_size = "larger", bold = T, align = "l") %>%
  kable_styling()
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
dir.create(file.path(getwd(), "DE_Results_2022/BetweenStages"))
tables <- lapply(contrasts.stage, function(contr) {
  tmp2 <- topTable(fitmm, coef = contr, n = Inf, sort.by = "P")
  tmp2$Gene.stable.ID <- rownames(tmp2)
  tmp2 <- left_join(tmp2, anno)
  tmp2 <- dplyr::select(tmp2, Gene.stable.ID, logFC, t, AveExpr, P.Value, adj.P.Val, DB.xref, Description, Ensembl, NCBI)
  filename <- gsub(" - ", "_v_", contr)
  filename <- gsub(", ", "_", filename, fixed = T)
  filename <- gsub("+", "", filename, fixed = T)
  filename <- gsub(" ", "_", filename, fixed = T)
  filename <- paste0(filename, "_DE_analysis")
  filename <- paste0(filename, ".txt")
  
  write.table(tmp2, file = paste0("./DE_Results_2022/BetweenStages/",filename), sep = "\t", quote = F, row.names = F)
  return(list('', nump = length(which(tmp2$adj.P.Val < 0.05))))
})

sigps <- unlist(lapply(tables, function(x)x[[2]]))
sumtab <- data.frame(Comparison = contrasts.stage, `Number of genes with adjusted P < 0.05` = sigps,
                     check.names = F)

sumtab %>%
  kbl(align = 'c') %>%
  add_header_above(c("Overview of results" = 2), font_size = "larger", bold = T, align = "l") %>%
  kable_styling()
```


### Session Info
```{r session}
sessionInfo()
```
